---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = F,
  warning = F,
  error = F
)
```

# DataForIrina

<!-- badges: start -->
<!-- badges: end -->

The goal of DataForIrina is to document the steps taken for the Dataset provided to Irina

## load needed packages

```{r loadPackages}
library(dplyr)
library(data.table)
library(readxl)
```


we will start by reading the dataset delivered to us from from the biodiversiy council, this dataset was generated on the 21st of September of 2022.

```{r, readDataset}
ToClean <- readxl::read_xlsx("2022-09-21.xlsx") 
```


# Filter data

We then filter the dataset to include only plants

```{r onlyPlants}
OnlyPlants <- ToClean |> 
  # make all names machine readable
  janitor::clean_names() |>
  # Filter to Kingdom Plantae
  dplyr::filter(rige == "Plantae")
```


Bot not all this have been resolved to species level, in the table bellow we see the number of taxa that have been resolved to which level

```{r, echo=FALSE}
OnlyPlants |> 
  dplyr::group_by(taxonrang) |> 
  dplyr::tally() |> 
  dplyr::arrange(desc(n)) |> 
  knitr::kable()
```


We want to keep only some of this, so we only include the ones that are to the level of Species, Form, Subspecies, and Variety (Art, Form, Underart, Varietet).

```{r}
OnlyPlants <- OnlyPlants |> 
  # Filter only some resolutions
  dplyr::filter(taxonrang %in% c("Art", "Form", "Underart", "Varietet")) |> 
# Ensure no duplecates
  dplyr::distinct()
```

```{r, echo=FALSE}
openxlsx::write.xlsx(OnlyPlants, "OnlyPlants.xlsx")
```


This leaves us with `r prettyNum(nrow(OnlyPlants), big.mark = ",")` valid taxa. this dataset can be downloaded [here](https://github.com/Sustainscapes/DataForIrina/raw/master/OnlyPlants.xlsx)

## Remove introduced species

Another dataset was made where we removed introduced species

```{r}
OnlyNatives <- OnlyPlants |> 
  dplyr::filter(herkomst != "Introduceret" | is.na(herkomst)) 
```


```{r, echo=FALSE}
openxlsx::write.xlsx(OnlyNatives, "OnlyNatives.xlsx")
```


This leaves us with `r prettyNum(nrow(OnlyNatives), big.mark = ",")` valid taxa. this dataset can be downloaded [here](https://github.com/Sustainscapes/DataForIrina/raw/master/OnlyNatives.xlsx)


# Download Datasets

- [OnlyPlants Dataset](https://github.com/Sustainscapes/DataForIrina/raw/master/OnlyPlants.xlsx): Filtered dataset including only plant taxa.

- [OnlyNatives Dataset](https://github.com/Sustainscapes/DataForIrina/raw/master/OnlyNatives.xlsx): Dataset with introduced species removed.
